Topic 6.2: Open Data

### 2020-04-22 draft_chapter.Rmd
finally knit stupid pdf

---
title: "Topic 6.2: Open Data"
author: "Samuel Langton (MMU) & Reka Solymosi (University of Manchester)"
date: "June 2020"
output: pdf_document
editor_options: 
  chunk_output_type: console
bibliography: refs.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
library(osmdata)
library(ggplot2)
library(sf)
library(readr)
library(dplyr)
library(tidyr)
```

@@

- **Availability and Access:** the data must be available in the publica domain, accessible at no more than a reasonable reproduction cost. It should be in a machine readable form that is easily accessed and modified. For example, it should be downloadable as a spreadsheet, or json file, rather than presented as a summary table in a PDF document. 
- **Re-use and Redistribution:** the data must be provided under terms that allow free use, including modification, separation, or compilation of the data. permit re-use and redistribution including the intermixing with other datasets.
- **Universal Participation:** everyone must be able to use, re-use and redistribute the data and its derivatives without restriction (eg: 'non-commercial' restrictions or use in education purposes only)

@@

Each one of the above categories might have its own strengths and limitations, and it is very important to consdier the source of the data you are working with, and what might mean for what and who is represented or excluded from these data. Here we will provide a general overview, and in the next section some guidance on how to make sure you take all necessary precautions when analysing open data of any kind.

Open data is hailed for its ease of use afforded by its availability and access. This strength is emphasised throughout the chapter. However another advantage and motivation for using open data is its potential to address many limitations associated with traditional surveying methods, such as social desirability bias, or issues associated with memory and recall [@mayer2013big].

However this also creates a limitation for such data, in that because it is already available, the researcher has no say into what gets collected, and instead must make do with what is available. There are limits as to what questions researchers can ask of a data set, and what interpretations are appropriate [@boyd2012critical].

Despite these limitation, the social sciences must embrace these new forms of data that, although messy, biased and noisy, have the potential to describe social phenomena better than well-organized small surveys or even national censuses [@savage2007coming].

## What do researchers need to know to make good use of open data

in the case of open data, there is no need to ask for permissions or negotiate data access in order to use it for research [@manovich2011trending]. In this sense, the great advantage of this data is that it is already out there. To collect it, all that is needed is a way to be able to interpret what this data means (for example by applying a framework for its analysis), and some skill in data scraping, wrangling and cleaning, in order to be able to transform it into a usable format for research [@boyd2012critical].

necessary to ask critical questions about what all this data means, who gets access to what data, how data analysis is deployed, and to what ends [@boyd2012critical].

be careful in interpreting people's communication online as authentic [@manovich2011trending].

The key concerns about these dara relate to issues about sampling, and ultimately the generalisability of any findings that can be drawn from them. In the case of crowdsourced data, where the sample of the population who contributes to such data is self-selected, giving way for people more motivated to speak about the issue, which means the data are likely to contain inherent bias [@longley2012geodemographics]. Specifically it tends to be men, between ages of 20-50, with a college or university degree who are most likely contributors [@budhathoki2010participants; @haklay2010good].

Looking into what contextual factors influence participation in Open Street Map, @mashhadi2013putting find that, socio-economic factors such as population density, dynamic population, distance from the centre and poverty all play an important role. These are important to keep in mind when reporting findings based on analysis of such data.

even with this limitation, the inclusion of crowdsourced data which are both up-to-date and specific to the problem at hand can still provide new insight in addition to the knowledge from established sources of data collected in traditional methods [@birkin2011calibration]. All of this helps us to understand the relations between the events and the occurrences that come together in unique places and provides a framework for understanding how places change over different time periods [@lonlongley2012geodemographics].

## Open Street Map

### What is Open Street Map?

- Brief history.
- Motivations for and advantages of an open web mapping platform.
- Primary features (keys, values, elements).
- Contributing to the data yourself.

### Downloading Open Street Map Data

- Define and explain APIs.
- For OSM, describe overpass queries and overpass-turbo.
- Query example (brief).

### Using Open Street Map data in R

- Made easy through the `osmdata` package in R.
- More straightforward than overpass-turbo.
- Compatible with both `sp` and `sf` classes of spatial data.
- We focus on `sf` as its compatible with the `tidyverse`.
- Outline the key documentation and references.

@@

First, ensure that you have the relevant packages installed in R. Although the main package used in this demonstration is `osmdata`, for querying the Open Street Map API, we use a number of additional packages for data handling and visualisation. If you don't have these packages installed, use the `install.packages()` function prior to loading each one with `library()`.

```{r, eval = F}
library(osmdata)
library(ggplot2)
library(sf)
library(readr)
library(dplyr)
library(tidyr)
```

>>>>

---
title: "Topic 6.2: Open Data"
author: "Samuel Langton (MMU) & Reka Solymosi (University of Manchester)"
date: "June 2020"
output:
  pdf_document:
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
bibliography: refs.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
library(jsonlite)
library(osmdata)
library(ggplot2)
library(sf)
library(readr)
library(dplyr)
library(tidyr)
```

@@

- **Availability and Access:** the data must be available in the publica domain, accessible at no more than a reasonable reproduction cost. It should be in a machine readable form that is easily accessed and modified. For example, it should be downloadable as a spreadsheet, or json file, rather than presented as a summary table in a PDF document. 
- **Re-use and Redistribution:** the data must be provided under terms that allow free use, including modification, separation, or compilation of the data. permit re-use and redistribution including the intermixing with other datasets.
- **Universal Participation:** everyone must be able to use, re-use and redistribute the data and its derivatives without restriction (eg: 'non-commercia' restrictions or use in education purposes only)

@@

Each one of the above categories might have its own strengths and limitations, and it is very important to consdier the source of the data you are working with, and what might mean for what and who is represented or excluded from these data. Here we will provide a general overview, and some guidance on how to make sure you take all necessary precautions when analysing open data of any kind.

Open data is hailed for its ease of use afforded by its availability and access; there is no need to ask for permissions or negotiate data access in order to use it for research [@manovich2011trending]. In this sense, the great advantage of this data is that it is already out there. To collect it, all that is needed is a way to be able to interpret what this data means (for example by applying a framework for its analysis), and some skill in data scraping, wrangling and cleaning, in order to be able to transform it into a usable format for research [@boyd2012critical].

But besides being easy to access, there are further advantages to using such data. One specific motivation for using open data comes from its potential to address many limitations associated with traditional surveying methods, such as social desirability bias, or issues associated with memory and recall [@mayer2013big]. By generating data often as a byproduct of other activities, we might gain more honest insight into people's everyday lives and associated social processes [@solymosi2018role]. At the same time, there are new threats to validity that may emerge in these sorts of data, and researchers should be careful in interpreting people's communication online as authentic [@manovich2011trending].

A related, and key concern about these dara relates to issues about sampling, and ultimately the generalisability of any findings that can be drawn from them. In the case of crowdsourced data, where the sample of the population who contributes to such data is self-selected, giving way for people more motivated to speak about the issue, which means the data are likely to contain inherent bias [@longley2012geodemographics]. Specifically it tends to be men, between ages of 20-50, with a college or university degree who are most likely contributors [@budhathoki2010participants; @haklay2010good].

Looking into what contextual factors influence participation in Open Street Map, @mashhadi2013putting found that, socio-economic factors such as population density, dynamic population, distance from the centre and poverty all play an important role. These are important to keep in mind when reporting findings based on analysis of such data.

Despite these limitation, the social sciences must embrace these new forms of data that, although messy, biased and noisy, have the potential to describe social phenomena better than well-organized small surveys or even national censuses [@savage2007coming]. The inclusion of crowdsourced data which are both up-to-date and specific to the problem at hand can still provide new insight in addition to the knowledge from established sources of data collected in traditional methods [@birkin2011calibration]. All of this helps us to understand the relations between the events and the occurrences that come together in unique places and provides a framework for understanding how places change over different time periods [@longley2012geodemographics].

---

In sum, researchers and analysts using open data must first ask critical questions about:

- **where** does this data come from? 
- **why** was the data collected in the first place?
- **who** is represented in the data, and who is excluded?
- **what** concepts and constructs can and cannot be operationalised with this data?
- **when** did data collection take place, and how might have that influenced results?

And meaningfully engage with the answers to these questions during conceptualisation and operationalisation phases of any research project.

---

# Practical exercise

In this exercise we will demonstrate:

- how to access open data using Application Programming Interfaces (APIs) and through downloading from official sites
- an example of how to clean and wrangle data for analysis
- how to link distinct data sets
- how open data from different sources may tell us different things, and why engaging with the critical questions outline above is crucial to properly use open data in research and analysis.

## Our aim

We are interested in finding out how bus stops might affect crime.

<!-- more about link between bus stops and crime here maybe something from Andy Newton/Vania Ceccato/ Anastasia Loukatou-Sideris/me -->

## Accessing data

We will be using two different types of open data in this exercise, public sector data (crime data and transport authority data) and crowdsourced data (from Open Street Map). To access them we will use three different methods: 1) Direct download, 2) Direct request to an API, 3) Request to an API using a wrapper. This will give you few different ideas about how you might go about accessing other open data sets relevant to your research.

### Direct download

The simplest way that open data is often made available is through direct download from a website. This is the case for open police data made available in the United Kingdom through the [data.police.uk](https://data.police.uk/) webportal. To access this data, simply type the url into any web browser ( [https://data.police.uk/](https://data.police.uk/) ), and follow the guidance to acquire data. In the case of police.uk you [select the tab which says "Data"](https://data.police.uk/data/), choose your "Date range:" (ie for what months you would like to download crime data for), choose the police forces you want crime data from, choose some options (whether you want to include crime data, outcomes data, and/or stop and search data) and click on 'Generate File' to download.

Once you download this data you will have to save this locally on your computer. We have saved ours in a subfolder called "data" and will be loading the data into R directly from there.

Some data downloads may come bundled with a data dictionary, explaining important information about your data set, which you will need to answer the critical questions outlined in previous sections. In other cases this might be found online. This is the case with the police.uk dataset, where the data dictionary can be found on the ["About" page](https://data.police.uk/about/). Make sure you always read through this correctly so you can answer your critical questions and fully understand your data set.

### Direct request to an API

An application programming interface (API) is a tool which defines an interface for a programme to interact with a software component or system, for example, what sort of requests or calls can be made to it, and how these can be made. We are specifically using the term APIs to denote tools created by an open data provider to provide access to different subsets of their content. APIs facilitate scripted and programmatic extraction of content, as permitted by the API provider [@olmedilla2016harvesting]. APIs can take many different forms and be of varying quality and usefulness [@foster2016big]. For the purposes of accessing open data from the web, we are specifically talking about **RESTful APIs**. REST stands for Representational State Transfer. These APIs work directly over the web, meaning that we as users can play with the API with relative ease to understand how it works [@foster2016big].

Here we will use the example of the British transit authority Transport for London (TfL) to practice constructing HTTP requests via their API, and then parsing the data that are returned. Much like how downloading data directly requires access to a data dictionary to make sense of the data and answer our critical questions, we must find a similar document to understand how the API we plan on using works. In the case of the TfL API this is provided via the [TfL Open Data page for their unified API](https://tfl.gov.uk/info-for/open-data-users/unified-api) and the [documentation page](https://api-portal.tfl.gov.uk/docs). We will engage more with this in the exercises.

### Request to an API using a wrapper

Often developers who work with APIs will share their code for doing so, and release them in the form of a package or module. This is called a **wrapper** because it's code that wraps the API in a nice package, making it easier for future users to access. The wrapper can take many forms, it could be a Python module, or an R package, or it can even be a web interface that provides a graphical user interface (GUI) for accessing the API in question.

To demonstrate this, we will be accessing data from Open Street Map (OSM). OpenStreetMap is a crowdsourced open data set of spatial information built by a community of mappers that contribute and maintain data about all sorts of environmental features, such as roads, trails, cafes, railway stations, and much more, all over the world. The result is a map rich in local knowledge, which prvides valuable information about places and their features.

OSM has two types of wrappers available for its API, a web-based GUI called [Overpass Turbo](https://overpass-turbo.eu/), and an R package called `osmdata`. We will explore using both.

<!-- ### Open Street Map -->

<!-- #### What is Open Street Map? -->

<!-- - Brief history. -->
<!-- - Motivations for and advantages of an open web mapping platform. -->
<!-- - Primary features (keys, values, elements). -->
<!-- - Contributing to the data yourself. -->

<!-- ### data.police.uk -->

<!-- ### Transport for London Open Data Portal -->

<!-- ### Downloading Open Street Map Data -->

<!-- - Define and explain APIs. -->
<!-- - For OSM, describe overpass queries and overpass-turbo. -->
<!-- - Query example (brief). -->

<!-- ### Using Open Street Map data in R -->

<!-- - Made easy through the `osmdata` package in R. -->
<!-- - More straightforward than overpass-turbo. -->
<!-- - Compatible with both `sp` and `sf` classes of spatial data. -->
<!-- - We focus on `sf` as its compatible with the `tidyverse`. -->
<!-- - Outline the key documentation and references. -->

@@

Let's get started! Getting back to our problem at hand, crime at bus stops, we want to know how crimes and bus stops interact...
<!-- clear up what research q we can actually answer....!! -->

First, ensure that you have the relevant packages installed in R. Although the main packages used in this demonstration are `rjson` for parsing the TfL API results and `osmdata` for querying the Open Street Map API, we use a number of additional packages for data handling and visualisation. If you don't have these packages installed, use the `install.packages()` function prior to loading each one with `library()`.

```{r, eval = F}
library(jsonlite)
library(osmdata)
library(ggplot2)
library(sf)
library(readr)
library(dplyr)
library(tidyr)
```

Now let's focus on acquiring our different data sets

### Direct Dowload of crime data

<!-- IDK how much detail we actually need here, probably not loads, about downloading extracting and importing police.uk data -->

### Direct request to TfL API

Our first step is to read through the . The tells us what sort of data is possible to acquire. We know that currently we want to find the location of bus stops, and any information that is available from TfL about these bus stops. In the documentation is where we find specifically *how* we can request this.

Specifically, we can scroll down to where we find "". Next to this we see the example URL: `https://api.tfl.gov.uk/line/24/route/sequence/outbound` This example demonstrates how we can request this information, and gives us a template to alter, but also allows us to take a peak into the result this call will return.

To see, try to take the above URL, and copy and paste it into your browser window. What do you see? It should be something like this:

![](img/tfl_api_browser.png)

This contains the information we requested (bus route 24 stops outbound) in a format called **JSON**, which stands for JavaScript Object Notation. JSON is an open standard way of storing and exchanging data, and will most likely be the format in which data are returned from most API calls.

Great, but how do we get this into R? Well this is where the `rjson` package is handy, and specifically we can use the `fromJSON()` function to parse this JSON p=object, which we can read directly from the URL into R using the `readLines()` function (base R). So, if we wanted the results of the bus route 24, we would type:

```{r getrt24}

api_call <- fromJSON(readLines("https://api.tfl.gov.uk/line/24/route/sequence/outbound"))

```

This gives us an object (`api_call`) which contains all the information returned by the TfL API. Now JSON is slightly different to dataframes which we are more used to dealing with, in that there is nested information

```{r conferttodf}

json_data_frame <- as.data.frame(api_call)

```

<!-- insert here tutorial on parsing the api data, this is a job for next week i'm tired -->

### Using wrappers to access Open Street Map data

<!-- Maybe start with the GUI-based one, and then do the below in R? -->
